The state of knowledge for marine invertebrate biodiversity in the continental US

### 2017-06-19 manuscript.Rmd
edit manuscript

The continental US includes a wide diversity of marine ecosystems from the Florida Keys, to the Gulf of Maine on the East Coast; and from the arid Southern California to the oceanic coast of the Pacific Northwest. With estuaries, bays, sandy beaches, rocky shores, and coral reefs, the US coastlines harbors many biotas. While the continental shelf extends for several hundred kilometers on the East coast, it is narrow on the West coast. This diversity of habitats and depths, combined with the exploitation of natural resources as early as the 1880, has led to extensive exploration and documentation of marine biodiversity [@Fautin2011]. However, apart from the synthesis that emerged from the Census of Marine Life efforts [@Fauntin2011], no study has attempted to synthesize the current state of knowledge of marine biodiversity in the US.

@@

To compare the faunistic lists from the Gulf of Mexico and the Pacific Northwest with records in public databases, we queried the iDigBio (https://portal.idigbio.org) and OBIS (http://www.iobis.org/) databases for all records they included in the same geographic boundaries as the lists. After filtering these lists to only include the same higher taxonomic level taxa as the species lists, and keeping only records for which the species could be matched to a species name in WoRMS, we had for the Gulf of Mexico `r format(nrow(idigbio_gom_records), big.mark=",")` and `r format(nrow(obis_gom_records), big.mark=",")` records for iDigBio and OBIS respectively, representing `r format(dplyr::n_distinct(idigbio_gom_records$worms_valid_name), big.mark=",")` and `r format(dplyr::n_distinct(obis_gom_records$worms_valid_name), big.mark=",")` species. For the Pacific Northwest, we retrieved `r format(nrow(idigbio_pnw_records), big.mark=",")` and `r format(nrow(obis_kozloff_records), big.mark=",")` records for iDigBio and OBIS respectively representing `r format(dplyr::n_distinct(idigbio_pnw_records$worms_valid_name), big.mark=",")` and `r format(dplyr::n_distinct(obis_kozloff_records$worms_valid_name), big.mark=",")` species. For the Gulf of Mexico, the majority of the species included in the expert lists are recorded in both iDigBio and OBIS (`r format_output(species_in_common$n_spp_gom_idigbio_obis/species_in_common$n_spp_gom*100)` %), or only OBIS (`r format_output(species_in_common$n_spp_gom_obis/species_in_common$n_spp_gom*100)` %) (Fig. \@ref(fig:database-overlap)). For the Pacific Northwest, these proportions were much lower with only `r format_output(species_in_common$n_spp_kozloff_idigbio_obis/species_in_common$n_spp_kozloff*100)`% of the species from the regional list with records in OBIS and iDigBio, and `r format_output(species_in_common$n_spp_kozloff_obis/species_in_common$n_spp_kozloff*100)`%  in OBIS only (Fig. \@ref(fig:database-coverage)). Previous esimates for species richness in the Gulf of Mexico for marine invertebrates were of `r format(11150-1541-434, big.mark=",")` species, and `r format(8374-909-134)` for the Pacific coast (including California) [@Fautin2010].

@@

```{r compare_database_overlap, fig.cap='Comparison of species by phylum across the public databases and the faunistic list. \\#fig:database-overlap', ech=FALSE}
knitr::include_graphics("figures/database_overlap.pdf")
```

@@

Increasing the computing literacy of the staff of museum in order for them to manipulate and use the data from their collection database is equally critical. As data across institutions get pooled into the large biodiversity information systems, it is crucial that data standards such as Darwin Core are embraced by the community.

Another potential venue to improve the quality of the data in biodiversity information systems, is better incitives for scientisits to publish their occurrence records alongside their papers as it is done for genetic data with GenBank [@Huang2011].

@@

This study highlights that we have too little data about too few species of marine invertebrates, we recommend that additional coordinated field expeditions are organized to sample more thoroughly the marine environment. The relation between sampling effort and diversity from museum collections makes clear that any additional samples will improve greatly our knowledge of marine biodiversity.

Samples collected during these expeditions need to be photographed to document live appearance that would otherwise be lost during preservation. We also need to capture metadata associated with the specimens to be able to understand better the physical characteristics of the habitats for these species. Specimens need to be sub-sampled to create material suitable for genomic analysis. Parts of data and specimen collection can be automated and captured digitally directly in the field speeding up data collection.

Coupling these more traditional approaches of biodiversity documentation with metabarcoding is a promising approach to screen an habitat for its potential for species that do not yet occur in public databases. <!-- explain why better --> Metabarcoding can complement and take advantage of the DNA barcodes already available. Additionally, in the future, data generated by metabarcoding efforts can be re-interpreted as more DNA barcodes associated with correctly identified specimens become available. Building the library of DNA barcodes in parallel of metabarcoding efforts ensures that matches in the database will correspond to actual specimens deposited in natural history collections; ruling out possible artifacts of the method. Information about species identity and distribution will thus build on existing natural history data rather than being limited to matching DNA sequences across studies.

Assembling species lists is a necessary but not sufficient step to understand biodiversity. Beyond counting the number of species, we need to understand better the geographical distribution of species, their abundance, and their biology. In order to predict the impact of climate change and other anthropogenic perturbations on species distributions, we need to collect more occurences for species. One potential challenge for this is to be able to convey taxonomic concepts for mostly unamed species. Developing robust identifiers that can be used across biodiversity information systems, and potentially anchored in results from DNA barcoding or other molecular techniques, is needed to collaboratively build our knowledge of species distributions.

Beyond their utility in documenting diversity, metabarcoding is also a promising technology to monitor changes in communities and for understanding the functioning of ecosystems. For instance, despite our limited knowledge of the diversity of communities in the meiofauna, metabarcoding could be used to detect and documents changes in species assemblages following the Deepwater Horizon Oil Spill [@Bik2012]. Metabarcoding has also been used to reveal important differences in the diet of 3 related species of fishes that co-occur [@Leray2015b], providing a much more detailed understanding of trophic relations.

However, taxonomists and experts are needed to collect, sort, process, and identify these samples and the data that come along with them. Even if technological advances can greatly speed up the documentation of species and their distributions, trained personnel is needed. Taxonomic expertise is dwindling, and there are now only a handful of US natural history museums with active marine invertebrates collections.

>>>>

The continental US includes a wide diversity of marine ecosystems from the Florida Keys, to the Gulf of Maine on the East Coast; and from the arid Southern California to the oceanic coast of the Pacific Northwest. With estuaries, bays, sandy beaches, rocky shores, and coral reefs, the US coastlines harbors many biotas. While the continental shelf extends for several hundred kilometers on the East coast, it is narrow on the West coast. This diversity of habitats and depths, combined with the exploitation of natural resources as early as the 1880, has led to extensive exploration and documentation of marine biodiversity [@Fautin2010]. However, apart from the synthesis that emerged from the Census of Marine Life efforts [@Fautin2010], no study has attempted to synthesize the current state of knowledge of marine biodiversity across the continental US.

@@

To compare the faunistic lists from the Gulf of Mexico and the Pacific Northwest with records in public databases, we queried the iDigBio (https://portal.idigbio.org) and OBIS (http://www.iobis.org/) databases for all records they included in the same geographic boundaries as the lists. After filtering these lists to only include the same higher taxonomic level taxa as the species lists, and keeping only records for which the species could be matched to a species name in WoRMS, we had for the Gulf of Mexico `r format(nrow(idigbio_gom_records), big.mark=",")` and `r format(nrow(obis_gom_records), big.mark=",")` records for iDigBio and OBIS respectively, representing `r format(dplyr::n_distinct(idigbio_gom_records$worms_valid_name), big.mark=",")` and `r format(dplyr::n_distinct(obis_gom_records$worms_valid_name), big.mark=",")` species. For the Pacific Northwest, we retrieved `r format(nrow(idigbio_pnw_records), big.mark=",")` and `r format(nrow(obis_kozloff_records), big.mark=",")` records for iDigBio and OBIS respectively representing `r format(dplyr::n_distinct(idigbio_pnw_records$worms_valid_name), big.mark=",")` and `r format(dplyr::n_distinct(obis_kozloff_records$worms_valid_name), big.mark=",")` species. For the Gulf of Mexico, the majority of the species included in the expert lists are recorded in both iDigBio and OBIS (`r format_output(species_in_common$n_spp_gom_idigbio_obis/species_in_common$n_spp_gom*100)` %), or only OBIS (`r format_output(species_in_common$n_spp_gom_obis/species_in_common$n_spp_gom*100)` %) (Fig. \@ref(fig:database-overlap)). For the Pacific Northwest, these proportions were much lower with only `r format_output(species_in_common$n_spp_kozloff_idigbio_obis/species_in_common$n_spp_kozloff*100)`% of the species from the regional list with records in OBIS and iDigBio, and `r format_output(species_in_common$n_spp_kozloff_obis/species_in_common$n_spp_kozloff*100)`%  in OBIS only (Fig. \@ref(fig:database-coverage)). Previous esimates for species richness in the Gulf of Mexico for marine invertebrates were of `r format(11150-1541-434, big.mark=",")` species, and `r format(8374-909-134, big.mark=",")` for the Pacific coast (including California) [@Fautin2010].

@@

```{r compare_database_overlap, fig.cap='Comparison of species by phylum across the public databases and the faunistic list. (\\#fig:database-overlap)', ech=FALSE}
knitr::include_graphics("figures/database_overlap.pdf")
```

@@

Increasing the computing literacy of the staff of museums in order for them to manipulate and use the data from their collection database is equally critical. As data across institutions get pooled into the large biodiversity information systems, it is crucial that data standards such as Darwin Core are embraced by the community.

Another potential venue to improve the quality of the data in biodiversity information systems, is better incitives from journal editors and funding agencies for scientists to publish their occurrence records in databases alongside their papers as it is done for genetic data with GenBank [@Huang2011e].

@@

This study highlights that we have too little data about too few species of marine invertebrates, we recommend that additional coordinated field expeditions are organized to sample more thoroughly the marine environment. The relation between sampling effort and diversity from museum collections makes clear that any additional samples will improve greatly our knowledge of marine biodiversity. Concerted and collaborative efforts are needed to share expertise and limited resources.

Samples collected during these expeditions need to be photographed to document live appearance that would otherwise be lost during preservation. We also need to capture metadata associated with the specimens to be able to understand better the physical characteristics of the habitats for these species. Specimens need to be sub-sampled to create material suitable for genomic analysis. Parts of data acquisition associated with specimens and their biota can be automated and captured digitally directly in the field, speeding up their ingestion in database collections.

Integrating metabarcoding as part of the data collection conducted in biodiversity surveys is a promising approach to screen an habitat for its potential to lead to new species or species that have not yet public DNA barcodes. For instance, by analyzing metabarcoding samples that contain planktonic larvae from a given habitat, we could get an indication of which taxonomic group would benefit most from additional sampling effort or give an indication of the proportion of species that has not been sampled yet. Additionally, data generated by metabarcoding efforts can be re-interpreted as more DNA barcodes associated with correctly identified specimens become available. Building the library of DNA barcodes in parallel of metabarcoding efforts ensures that database matches correspond to actual specimens deposited in natural history collections, ruling out possible artifacts of the method. With such approaches, the geographic distribution of species can build on and linked to existing records rather than being limited to matching DNA sequences across studies.

Assembling species lists is a necessary but not sufficient step in our quest to understand biodiversity. We also need to understand better the geographic distribution of biodiversity, how the abundance of species vary in space and time, and their role in the functionning of ecosystems. In order to predict the impact of climate change and other anthropogenic perturbations on species distributions, we need to collect additional occurences for species as well as the characteristics of their habitats. Among the challenges for this enterprise, we need to be able to convey taxonomic concepts for species which, for the most part, have not formally been named yet. Developing robust identifiers that can be shared across biodiversity information systems, and anchored in results from DNA barcoding or other molecular techniques, is needed to collaboratively build our knowledge of species distributions.

Metabarcoding also offers to be a promising technology to monitor changes in communities and for understanding the functioning of ecosystems. For instance, despite our limited knowledge of the diversity of communities in the meiofauna, metabarcoding has been successfully used to detect and documents changes in species assemblages following the Deepwater Horizon Oil Spill [@Bik2012]. Metabarcoding has also revealed important differences in the diet of 3 related species of fishes that co-occur at small spatial scales, providing a much more detailed understanding of trophic relations [@Leray2015b].

Even if technological advances can greatly speed up the documentation of species and their distributions, taxonomists and experts are needed  to collect, sort, process, and identify samples. Taxonomic expertise within the US is dwindling, and there are now only a handful of US natural history museums with active marine invertebrates collections, and curatorial staff.
